Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)

### 2017-04-08 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor modifications

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% c('MSSM_RNA_ACC_BP_24', 
                                          'MSSM_RNA_ACC_375'))] = ''

@@

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375')),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
p

indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375')
indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals

@@

@@

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```

>>>>

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
indToRemove = c('MSSM_RNA_ACC_BP_24',

                'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 
                'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220', 'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133',

                'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 
                'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56')

# Outliers from Menahce's DLPFC analysis syn3493755
# outliers = c('MSSM_RNA_BP_PFC_1', 'MSSM_RNA_PFC_21', 'MSSM_RNA_BP_PFC_3', 'MSSM_RNA_PFC_103', 'MSSM_RNA_PFC_105', 'MSSM_RNA_PFC_115', 'MSSM_RNA_PFC_169',  'MSSM_RNA_PFC_17',
#   'MSSM_RNA_PFC_18', 'MSSM_RNA_PFC_19', 'MSSM_RNA_PFC_213', 'MSSM_RNA_PFC_214', 'MSSM_RNA_PFC_215', 'MSSM_RNA_PFC_216', 'MSSM_RNA_PFC_217', 'MSSM_RNA_PFC_218', 
#   'MSSM_RNA_PFC_219', 'MSSM_RNA_PFC_222', 'MSSM_RNA_PFC_24', 'MSSM_RNA_PFC_252', 'MSSM_RNA_PFC_259', 'MSSM_RNA_PFC_263', 'MSSM_RNA_PFC_267', 'MSSM_RNA_PFC_275',
#   'MSSM_RNA_PFC_307', 'MSSM_RNA_PFC_322', 'MSSM_RNA_PFC_325', 'MSSM_RNA_PFC_27', 'MSSM_RNA_PFC_332', 'MSSM_RNA_PFC_333', 'MSSM_RNA_PFC_334', 'MSSM_RNA_PFC_335',
#   'MSSM_RNA_PFC_340', 'MSSM_RNA_PFC_35', 'MSSM_RNA_PFC_362', 'MSSM_RNA_PFC_372', 'MSSM_RNA_PFC_373', 'MSSM_RNA_PFC_385', 'MSSM_RNA_PFC_393', 'MSSM_RNA_PFC_395',
#   'MSSM_RNA_PFC_49', 'MSSM_RNA_PFC_54', 'MSSM_RNA_PFC_70', 'MSSM_RNA_PFC_76', 'MSSM_RNA_PFC_77', 'MSSM_RNA_PFC_78', 'MSSM_RNA_PFC_88', 'MSSM_RNA_PFC_93',
#   'FAKE_DLPFC_CMC_MSSM_426', 'PENN_RNA_PFC_29', 'FAKE_DLPFC_CMC_PENN_055', 'PENN_RNA_PFC_75', 'PENN_RNA_PFC_110', 'FAKE_DLPFC_CMC_PENN_106', 'FAKE_DLPFC_CMC_PENN_107',
#   'FAKE_DLPFC_CMC_PENN_108', 'MSSM_RNA_PFC_14', 'MSSM_RNA_BP_PFC_21', 'MSSM_RNA_PFC_293', 'MSSM_RNA_PFC_349', 'MSSM_RNA_PFC_353', 'MSSM_RNA_PFC_363', 'MSSM_RNA_PFC_369',
#   'PENN_RNA_PFC_82', 'PENN_RNA_PFC_89', 'PITT_RNA_PFC_1159', 'PITT_RNA_PFC_1296', 'PITT_RNA_PFC_537',  'PITT_RNA_PFC_567', 'PITT_RNA_PFC_871', 'PITT_RNA_PFC_878',
#   'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_210',  'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_278', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_361',
#   'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_56', 'PITT_RNA_BP_PFC_10003', 'PITT_RNA_BP_PFC_1047', 'PITT_RNA_BP_PFC_1086', 'PITT_RNA_BP_PFC_1092', 'PITT_RNA_BP_PFC_1247',
#   'PITT_RNA_BP_PFC_1282', 'PITT_RNA_BP_PFC_1324', 'PITT_RNA_BP_PFC_1374', 'PITT_RNA_BP_PFC_1391', 'PITT_RNA_BP_PFC_686')

# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''

@@

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% indToRemove),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
p

indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
COVARIATES = COVARIATES[indToRetain,]

tmp = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
  group_by(Dx, Tissue) %>%
  summarise(count = n()) %>%
  spread(Tissue, count)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals

@@

Distribution of samples are:
`r kable(tmp)`

@@

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
p
```
